---
title: "STAN"
author: "LAURA PAAABY"
date: "4/6/2022"
output: html_document
---
# STAN 

```{r}
pacman::p_load(rstan, rethinking)
```


THIS IS AN INTRO TO THE PURE STAN APPROACH 
- ulam just buils Stan code 
- Stan code is portable, runs on anything
- if you learn Stan, it can be used in any scripting language. 


okay this is a way to make stan codes - so the entire code below should be made in a separate stan script, which can then be pulled into the R document from files (in the block in the bottom left), and attached as a model as done in the bottom. 
You would thus NOT work this kind of way often. 



**FIRST EXAMPLE OF STAN CODE**
we would never work like this, cause this should all go into a STAN script file, that we should then pull it back to r by the file path as done in the bottom. 
```{r}
#### stan code (mHMC), this is an example using the variables from the divorce data.
data{
  ## the observed variables 
  vector[50] D; # divorce rate
  vector[50] A; # age 
  vector[50] M; # marriage
}

### ^^^^  Must declare the type of each observed variable so Stan can catch errors and know operations are allowed.
#So if you try to do something that is not allowed by the variable type

# UNOBSERVED VARIABLES 
parameters {
  # the unobserved variables 
  real a; #alpha
  real bM; # beta marriage
  real bA; #beta age
  real<lower=0> sigma; #sigma
}

### ^^^^^ the unobserved variables also need checks and constraints. Declared here.

# MODEL BLOCK
model{
  # compute the log posterior probability 
  vector[50], mu;
  sigma ~ exp(1);
  bA ~ dnorm(0,0.5);
  bM ~ dnorm(0,0.5);
  a ~ dnorm(0,0.2);
  for ( i in 1:50) {
    mu[i] = a + bM*M[i] + bA*A[i];
  }
D ~ normal(mu, sigma);
}

```

Declare the distribution parts of the model sufficient to compute *posterior probability*
In big models, this part is super complex.



#### SAVE STAN CODE AS OWN FILE 
apparently this is how it works using pure stan
=> to be able to sample posteriors we must make the code into a file:
```{r}
mHMC_stan <- (file = "Name_of_stan", data = dat) # obs => im not completely sure how this works

### extracting posteriors: 
post <- extract.samples(mHMC_stan)
```


















